---
title: "R Notebook"
output: html_notebook
---

```{r setup}
library(dmlpm)
library(tidyverse)
library(Rfast)
library(foreach)
library(vegan)
library(ggiraph)
```

# Example 1: n = 25, T = 25, K = 5

Generate data:

```{r fig.height=4, fig.width=12}
# Generate a random Z matrix:
set.seed(4)
d <- 2 # no. of latent dimensions
n <- 25 # nodes
K <- 5 # Layers
TT <- 25 # T time points

# Initialise the latent positions, Z_1, which are the same for each layer:
sigma2_0 <- 25 # the variance for the initial positions (tau in Sewell thesis)
Z <- array(NA, dim = c(n, 2, TT))
for(i in 1:n) 
  for(d in 1:2) 
    Z[i, d, 1] = rnorm(1, 0, sd = sqrt(sigma2_0))

# The remaining time points (1 < t <= T) are updated with Markovian dependence (again, same for each layer):
sigma2_z <- 0.25^2 # the amount of perturbation (SD) to give to the previous time point (sigma in Sewell thesis)
for(t in 2:TT)
  for(i in 1:n)
    for(d in 1:2)
      Z[i, d, t] = rnorm(n = 1, mean = Z[i, d, t-1], sd = sqrt(sigma2_z))


# Dynamic alpha (use a random walk to generate the alphas):
sigma2_alpha <- 0.25^2
alpha <- matrix(NA, nrow = K, ncol = TT)
alpha[ , 1] <- round(runif(K, 5, 10)) # initial values (t = 1)
for(t in 2:TT) {
  alpha[ , t] <- rnorm(K, mean = alpha[ , t - 1], sd = sqrt(sigma2_alpha))
}
#alpha
# Generate LPM and K x T adjacency matrices, one for each layer and time point:

#Y <- map(1:TT, function(t) LPM_gen(Z[ , , t], alpha)) %>% simplify2array() # maps are too confusing here, switch to for loops over the array
Y <- array(NA, dim = c(n, n, TT, K))
for(k in 1:K) {
  print(paste0("k = ", k))
  for(t in 1:TT) {
    Y[ , , t, k] <- LPM_gen_poisson(Z[ , , t], alpha[k, t])
  }
}

Y_tk_list <- map(1:TT, function(t) map(1:K, function(k) upper_tri(Y[ , , t, k])))
Y_list <- lapply(seq(dim(Y)[4]), function(k) Y[ , , , k]) # for Rcpp function
```

I can plot the network separately for each layer for now:

```{r fig.height=6, fig.width=12}
for(k in 1:K)
  print(plot_lpm_dynamic(Y = Y[ , , , k], Z = Z, alpha = alpha[k], center = T) + ggtitle(paste0("Layer ", k)))
```

I can also plot for each time point:

```{r fig.height=4, fig.width=12}
for(t in 1:TT)
  print(plot_lpm_multiview(Y = Y[ , , t, ], Z = Z[ , , t], alpha = alpha, center = T) + ggtitle(paste0("t = ", t)))
```


Plot the paths (same for each layer):

```{r}
print(plot_dynamic_path(Z) + ggtitle("Path of nodes over time"))
```

## Fit

R: Run time: 26.16 mins, n = 25, T = 6, K = 3, 4 chains
Rcpp: Run time: 3.13 mins, n = 25, T = 6, K = 3, 4 chains

Laptop comparing the declaration of arma objects vs straight to row:
Straight to row:   "Run time: 4.38 mins"
With declarations: "Run time: 4.54 mins" 

PC comparing the declaration of arma objects vs straight to row:
Straight to row:   "Run time: 2.3 mins"
With declarations: "Run time: 3.03 mins"


```{r}
#fit <- readRDS(file = "obj/n25_T6_K3_example_2_1")

# Initialise:
# (Rotate with reference to the first time point so that they are all in a "similar" orientation)
# Just use the first layer for now. Later I can try taking the sum across layers or something else.
Z_0 <- array(0, dim = c(n, 2, dim(Z)[3]))
Z_0[ , , 1] <- mds_initialise(Y[ , , 1, 1]) # t = 1, K = 1
for(t in 2:TT) {                            # t > 1
  Z_0[ , , t] <- vegan::procrustes(Z_0[ , , 1], mds_initialise(Y[ , , t, 1]))$Yrot
}

fit <- DMLPM(
  num_samples = 1e3, burn = 1e4,
  thin = 10, post_pred_thin_factor = 1e6, # large ppthin to avoid doing it, will fix later, it's too slow
  num_chains = 1,
  Y = Y,
  sigma2_0 = sigma2_0, sigma2_a = 100, sigma2_z = 0.01^2, 
  alpha_0 = matrix(1, nrow = K, ncol = TT), Z_0 = Z_0, #array(0, dim = c(n, 2, dim(Z)[3])),
  eta_alpha = 0.1, eta_z = 0.1,
  delta_alpha = 0.5, delta_z = 0.5,
  discard = TRUE
)

# Cache result:
#saveRDS(object = fit, file = "obj/n25_T6_K3_example_2_1")
```



```{r}
#fit <- readRDS(file = "obj/n25_T6_K3_example_2_1")

# Initialise:
# (Rotate with reference to the first time point so that they are all in a "similar" orientation)
# Just use the first layer for now. Later I can try taking the sum across layers or something else.
Z_0 <- array(0, dim = c(n, 2, dim(Z)[3]))
Z_0[ , , 1] <- mds_initialise(Y[ , , 1, 1]) # t = 1, K = 1
for(t in 2:TT) {                            # t > 1
  Z_0[ , , t] <- vegan::procrustes(Z_0[ , , 1], mds_initialise(Y[ , , t, 1]))$Yrot
}

fit <- MH_dynamic_multiview(
  num_samples = 1e3, burn = 5e4,
  thin = 10, post_pred_thin_factor = 10,
  num_chains = 4,
  Y = Y,
  sigma2_0 = sigma2_0, sigma2_a = 1, sigma2_z = 0.001, 
  alpha_0 = matrix(1, nrow = K, ncol = TT), Z_0 = Z_0, #array(0, dim = c(n, 2, dim(Z)[3])),
  eta_alpha = 0.1, eta_z = 0.1,
  delta_alpha = 0.5, delta_z = 0.5,
  discard = TRUE
)

# Cache result:
#saveRDS(object = fit, file = "obj/n25_T6_K3_example_2_1")
```

Check just out of interest:

```{r}
walk(1:TT, function(t) print(paste0("t = ", t, ": ", round(vegan::protest(Z[ , , t], Z_0[ , , t])$t0, 3))))
```


## Post processing


### Log-posterior

```{r echo=FALSE, fig.height=6, fig.width=12}
chain <- which.max(fit$map$i_map$log_posterior_max)
chain
df <- data.frame(i = 1:fit$num_samples, log_posterior = fit$mcmc_chains[[chain]]$log_posterior_thin)
log_posterior_true <- log_posterior(Y = Y, Z = Z, alpha = alpha, sigma2_0 = sigma2_0, sigma2_a = 100, sigma2_z = sigma2_z)
p1 <- ggplot(df) +
  geom_line(aes(i, log_posterior), colour = "steelblue", linewidth = 0.5, alpha = 0.8, show.legend = F) +
  labs(x = "i", y = "Log-posterior",
       title = paste0("Log-posterior after each (full) iteration (num_samples = ", fit$num_samples, ", thin = ", fit$thin, ")"), subtitle = paste0("Log posterior using true values = ", round(log_posterior_true, 1), " (black line)")) +
  #geom_hline(yintercept = log_posterior_true, col = "black", lty = 2) +
  theme(plot.title = element_text(size = 12))
p2 <- ggAcf2(fit$mcmc_chains[[chain]]$log_posterior_thin) +
  ggtitle("ACF of log-posterior")
p1 + p2
mean(fit$mcmc_chains[[chain]]$log_posterior_thin)
plot_log_posterior(fit, scales = "free")
```

Need to update the Rhat functions first to use the plot_alpha_posterior function

```{r}
# fit$K <- K
# fit$alpha_hat <- alpha_hat
# plot_alpha_posterior(fit, facet_trace = T)
```

```{r}
alpha-
apply(fit$mcmc_chains[[chain]]$alpha_out_thin, c(1, 2), mean)
```

```{r}
bind_rows(list(true = data.frame(cbind(k = c(1:K), alpha)), estimated = data.frame(cbind(k = 1:K, apply(fit$mcmc_chains[[chain]]$alpha_out_thin, c(1, 2), mean)))), .id = "set") %>% 
  pivot_longer(cols = starts_with("V"), names_to = "t", values_to = "alpha_kt") %>% 
  mutate(t = as.numeric(str_replace(t, "V", "")) - 1) %>% 
  mutate(set_k = str_c(set, k)) %>% 
  ggplot() +
  geom_line(aes(t, alpha_kt, group = set_k, colour = set)) +
  facet_wrap(~k, scales = "fixed")
```


Do each layer manually:
(a bit sticky but good posterior estimated of true values)


```{r echo=FALSE, fig.height=4, fig.width=10}
for(k in 1:K) {
  for(t in 1:TT) {
      df <- data.frame(i = 1:fit$num_samples, alpha = fit$mcmc_chains[[chain]]$alpha_out_thin[k, t, ])
    p1 <- ggplot(df) +
      geom_line(aes(i, alpha), colour = "steelblue", linewidth = 0.5, alpha = 0.8, show.legend = F) +
      geom_hline(yintercept = alpha[k, t], colour = "black", linetype = 1) + # true
      geom_hline(yintercept = fit$alpha_hat[k, t], colour = "black", linetype = 2) + # posterior mean
      labs(x = "i", y = "alpha",
           title = paste0("True alpha: ", round(alpha[k, t], 2), "; posterior mean: ", round(apply(fit$mcmc_chains[[chain]]$alpha_out_thin, c(1, 2), mean)[k, t], 2)))
    
    p2 <- ggAcf2(fit$mcmc_chains[[chain]]$alpha_out_thin[k, t, ]) +
      ggtitle("ACF of alpha")
    print(p1 + p2)
  }
}
```

## Acceptance rates

alpha:

```{r}
fit$mcmc_chains[[chain]]$alpha_acc_mean
fit$mcmc_chains[[chain]]$eta_alpha_i
```

$Z_t$:

```{r}
fit$mcmc_chains[[chain]]$Z_acc_mean %>% round(2)
```

## Estimated latent positions

Procrustes correlation for each time point:

```{r}
walk(1:TT, function(t) print(paste0("t = ", t, ": ", round(vegan::protest(Z[ , , t], fit$Z_hat_t[ , , t])$t0, 3))))

# Procrustes correlation of the stacked matrices:
# (first I need to take the posterior mean of the stacked and transformed samples, the Z_trans matrix was already made above)
Z_trans_mean <- matrix(NA, nrow = dim(fit$Z_trans)[1], ncol = dim(fit$Z_trans)[2])
for(i in 1:(n*TT)) {
  for(j in 1:2) {
    Z_trans_mean[i, j] <- mean(fit$Z_trans[i, j, ])
  }
}

# Construct stacked matrix of the true Z matrices:
Z_stack_true <- matrix(NA, nrow = n*TT, ncol = d)
for(t in 1:TT) {
  Z_stack_true[((t - 1)*n + 1):(t*n), ] <- Z[ , , t] # These indexes look weird but they are correct:
}

# Compare to the stacked true matrices:
paste0("Correlation of stacked matrices (nT x 2): ", round(vegan::protest(Z_stack_true, Z_trans_mean)$t0, 3))
```

This is the true node trajectories compared to the posterior estimates after
Procrustes transformation using the true values as reference:
(the same "stacking" method is used again here)

```{r echo=FALSE, fig.height=6, fig.width=12}
# Construct the nT x 2 reference matrix of all the stacked (true) Z_t matrices:
Z_ref <- matrix(NA, nrow = n*TT, ncol = 2)
for(t in 1:TT) {
  Z_ref[((t - 1)*n + 1):(t*n), ] <- Z[ , , t] # These indexes look weird but they are correct:
}

# Transform:
# I originally used Z_stack here but I think it's fine to use _trans again here since they are just a rotation of Z_stack anyway. I haven't returned Z_stack so I will just use Z_trans.
Z_trans <- array(NA, dim = dim(fit$Z_trans)) # Array to store the transformed samples dim = (nT, 2, num_samples)
for(i in 1:dim(Z_trans)[3]) {
  Z_trans[ , , i] <- procrustes(X = Z_ref, Y = fit$Z_trans[ , , i], scale = FALSE)$Yrot
}

# Unstack the matrices again:
Z_trans_t <- array(NA, dim = dim(fit$mcmc_chains[[chain]]$Z_out_thin))
for(t in 1:TT) {
  Z_trans_t[ , , t, ] <- Z_trans[((t - 1)*n + 1):(t*n), , ]
}

# Posterior means for each Zt:
Z_hat_t2 <- array(NA, dim = dim(Z_trans_t[ , , , 1]))
for(t in 1:TT) {
  for(i in 1:n) {
    for(j in 1:2) {
      Z_hat_t2[i, j, t] <- mean(Z_trans_t[i, j, t, ])
    }
  }
}

df1 <- plyr::alply(Z, .margins = 3) %>% # https://stackoverflow.com/a/20199008
  map(data.frame) %>% 
  bind_rows(.id = "t") %>% 
  mutate(t = as.numeric(t), i = rep(1:n, dim(Z)[3])) %>% 
  set_names(c("t", "z1", "z2", "i")) %>% 
  mutate(set = "true")

df2 <- plyr::alply(Z_hat_t2, .margins = 3) %>% # https://stackoverflow.com/a/20199008
  map(data.frame) %>% 
  bind_rows(.id = "t") %>% 
  mutate(t = as.numeric(t), i = rep(1:n, dim(Z)[3])) %>% 
  set_names(c("t", "z1", "z2", "i")) %>% 
  mutate(set = "estimated")

p <- rbind(df1, df2) %>% 
  mutate(set = factor(set, levels = c("true", "estimated"))) %>% 
  ggplot() +
  geom_path_interactive(aes(z1, z2, group = i, colour = t, tooltip = paste0("node ", i), data_id = paste0("node ", i)),
                        hover_css = "stroke:indianred;stroke-width:2", 
                        arrow = arrow(length = unit(1.5, "mm"))) +
  facet_wrap(~set, )
girafe(ggobj = p, width_svg = 12, height_svg = 6) %>% 
  girafe_options(opacity = 0.2)
```

Compare estimated latent positions at each time point:
(the positions have already been rotated with reference to the true positions)

TODO:
Add the comaprison of time points and layers using the code above where I generate the data
and this code for the comparison.

```{r echo=FALSE, fig.height=6, fig.width=12}
(plot_lpm_dynamic(Y, Z, alpha) + ggtitle("True")) + (plot_lpm_dynamic(Y, Z_hat_t2, round(alpha_hat, 2)) + ggtitle("Estimated"))
```


# Example 2: using Sewell init


```{r}
library(dmlpm)
Z_0_sewell <- initialise_Z(Y, d = 2)
```


```{r echo=FALSE, fig.height=6, fig.width=8}
Z_0_sewell <- initialise_Z(Y, d = 2)

# Construct the nT x 2 reference matrix of all the stacked (true) Z_t matrices:
Z_ref <- matrix(NA, nrow = n*TT, ncol = 2)
for(t in 1:TT) {
  Z_ref[((t - 1)*n + 1):(t*n), ] <- Z[ , , t] # These indexes look weird but they are correct:
}

Z_stack_sewell <- matrix(NA, nrow = n*TT, ncol = 2)
for(t in 1:TT) {
  Z_stack_sewell[((t - 1)*n + 1):(t*n), ] <- Z_0_sewell[ , , t] # These indexes look weird but they are correct:
}
Z_stack_sewell <- procrustes(X = Z_ref, Y = Z_stack_sewell, scale = FALSE)$Yrot
# Unstack:
Z_0_sewell_rotated <- array(NA, dim = dim(Z_0_sewell))
for(t in 1:TT) {
  Z_0_sewell_rotated[ , , t] <- Z_stack_sewell[((t - 1)*n + 1):(t*n), ] 
}


df1 <- plyr::alply(Z, .margins = 3) %>% # https://stackoverflow.com/a/20199008
  map(data.frame) %>% 
  bind_rows(.id = "t") %>% 
  mutate(t = as.numeric(t), i = rep(1:n, dim(Z)[3])) %>% 
  set_names(c("t", "z1", "z2", "i")) %>% 
  mutate(set = "true")

df2 <- plyr::alply(Z_0_sewell_rotated, .margins = 3) %>% # https://stackoverflow.com/a/20199008
  map(data.frame) %>% 
  bind_rows(.id = "t") %>% 
  mutate(t = as.numeric(t), i = rep(1:n, dim(Z)[3])) %>% 
  set_names(c("t", "z1", "z2", "i")) %>% 
  mutate(set = "estimated")

p <- rbind(df1, df2) %>% 
  mutate(set = factor(set, levels = c("true", "estimated"))) %>% 
  ggplot() +
  geom_path_interactive(aes(z1, z2, group = i, colour = t, tooltip = paste0("node ", i), data_id = paste0("node ", i)),
                        hover_css = "stroke:indianred;stroke-width:2", 
                        arrow = arrow(length = unit(1.5, "mm"))) +
  facet_wrap(~set, )
girafe(ggobj = p, width_svg = 12, height_svg = 6) %>% 
  girafe_options(opacity = 0.2)


for(t in 1:TT) {
  print((vegan::protest(Z_0_sewell[ , , t], Z[ , , t])$t0) %>% round(3))
}


alpha_0 <- initialise_alpha(Y, Z_0_sewell)

# Compare using plot from the the posterior comparioson below:
bind_rows(list(true = data.frame(cbind(k = c(1:K), alpha)), initial = data.frame(cbind(k = 1:K, alpha_0))), .id = "set") %>% 
  pivot_longer(cols = starts_with("V"), names_to = "t", values_to = "alpha_kt") %>% 
  mutate(t = as.numeric(str_replace(t, "V", "")) - 1) %>% 
  mutate(set_k = str_c(set, k)) %>% 
  ggplot() +
  geom_line(aes(t, alpha_kt, group = set_k, colour = set)) +
  facet_wrap(~k, scales = "fixed")
```
